Technique for \earlywarning\ detection of compact binary coalescence and its implications for multi-messenger astronomy

### 2011-06-06 conclusions.tex
lloid_paper: addressed comments from Steve and Missy; significant introduction reordering
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@944 ccdd05d5-56c1-461a-bf98-85f6876523b4

\editorial{We should drive home the point that our method is as fast as the fft convolution but without any latency at all.}%
We have demonstrated a computationally feasible procedure for the rapid and even preemptive
detection of gravitational waves emitted during the coalescence of neutron
stars and stellar-mass black holes. Our method is as fast as standard \fft{}
convolutions but without any latency at all.
%
\editorial{Drew: ``but without any latency at all" is awkward and should be reworded.}
%
Our search targets are expected to
produce prompt electromagnetic signals and may be the progenitors of some short
hard gamma-ray bursts.  Rapid alerts to the broader astronomical community will
improve the chances of detecting an electromagnetic counterpart in bands from
gamma-rays down to radio.  We anticipate requiring $\sim$$10\times4$ modern
multi-core computers to analyze a four-detector network of gravitational-wave data
for binary neutron stars and stellar-mass black holes.  This is well within the
current computing capabilities of the \textsc{lsc} Data Grid~\cite{LDG}.

>>>>

We have demonstrated a computationally feasible procedure for the rapid and even preemptive
detection of gravitational waves emitted during the coalescence of neutron
stars and stellar-mass black holes. Our method is as fast as standard \fft{}
convolutions but allows for sample-in-sample-out processing.
Our search targets are expected to
produce prompt electromagnetic signals and may be the progenitors of some short
hard gamma-ray bursts.  Rapid alerts to the broader astronomical community will
improve the chances of detecting an electromagnetic counterpart in bands from
gamma-rays down to radio.  We anticipate requiring $\sim$$10\times4$ modern
multi-core computers to analyze a four-detector network of gravitational-wave data
for binary neutron stars and stellar-mass black holes.  This is well within the
current computing capabilities of the \LIGO{} Data Grid~\cite{LDG}.

### 2011-06-06 inspiral_svd.tex
lloid_paper: addressed comments from Steve and Missy; significant introduction reordering
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@944 ccdd05d5-56c1-461a-bf98-85f6876523b4

\begin{abstract}
The rapid detection of compact binary coalescence with a network of advanced
gravitational-wave detectors will offer a unique opportunity for
multi-messenger astronomy.  Prompt detection alerts to the astronomical
community may make it possible to observe the onset of electromagnetic emission
from compact binary coalescence.  We demonstrate a computationally practical
analysis strategy that produces early warning triggers even before
gravitational radiation from the final merger has arrived at the detectors.
With current rate estimates for the Advanced \LIGO{} design configuration, it
should be possible to detect $\sim$10 sources earlier than 10 seconds before
merger in 1 year of live time using the method we describe.
\end{abstract}

>>>>

\begin{abstract}
The rapid detection of compact binary coalescence with a network of advanced
gravitational-wave detectors will offer a unique opportunity for
multi-messenger astronomy.  Prompt detection alerts to the astronomical
community may make it possible to observe the onset of electromagnetic emission
from compact binary coalescence.  We demonstrate a computationally practical
analysis strategy that produces early-warning triggers even before
gravitational radiation from the final merger has arrived at the detectors.
With current rate estimates for the Advanced \LIGO{} design configuration, it
should be possible to detect $\sim$10 such sources earlier than 10 seconds
before merger in 1 year of live time using the method we describe.
\end{abstract}

### 2011-06-06 introduction.tex
lloid_paper: addressed comments from Steve and Missy; significant introduction reordering
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@944 ccdd05d5-56c1-461a-bf98-85f6876523b4

\editorial{Would the introduction be more effective without this paragraph?, should we find whitepapers?}

The coalescence of compact binary systems consisting of neutron stars (\textsc{ns})
and/or black holes (\textsc{bh}) is the most promising source of gravitational radiation
for Advanced \LIGO, Virgo, \GEO\ and \LCGT~\cite{ALIGOWeb, AVirgoWeb, GEOWeb,
LCGTWeb}.  Tens of binary coalescence events are expected to be observed in the
advanced detector era later this decade~\cite{Abadie:2010p10836}.

As a compact binary system loses energy to gravitational waves, its orbital
separation decreases. This causes a run-away inspiral with the
gravitational-wave amplitude and frequency increasing until the system
eventually merges near the innermost stable circular orbit (\ISCO).  If a
neutron star is involved it may become tidally disrupted near the merger.  This
disrupted matter can fuel a bright electromagnetic counterpart in the system's
final moments as a binary~\cite{shibata:2007}.

Prompt electromagnetic emission can arise as shells of relativistically
out-flowing matter collide in the inner shock. Such an inner shock from a
compact binary coalescence is believed to be a mechanism for short gamma-ray
bursts (short \GRB{}s)~\cite{Lee:2005, nakar07}. The same inner shocks, or
potentially reverse shocks, can produce a bright accompanying optical flash~\cite{Sari99}. Prompt emission is a probe into the extreme
initial conditions of the outflow, in contrast with afterglows, which are
relatively insensitive to initial conditions. Optical flashes
have only been observed for a handful of long
\GRB{}s~\cite{2011CRPhy..12..255A}
by telescopes with extremely rapid response or, in the case of \textsc{grb 080319b}, by
pure serendipity, where several telescopes were observing a previous \GRB{} in the
same field~\cite{2008Natur.455..183R}. Short \GRB{}s, on the other hand, typically
fade too quickly to observe anything before the initial burst of gamma-rays and hard
x-rays. Rapid gravitational-wave transient alerts could enable the observation
of optical flashes from short \GRB{}s and perhaps even
precursors~\cite{0004-637X-723-2-1711}. An electromagnetic counterpart would
vastly boost
confidence in the gravitational-wave detection and provide the tight sky
localization necessary to allow determination of the source's host galaxy,
which leads to a redshift measurement. With both redshift and a coincident
gravitational-wave observation, we can produce precision measurements of the
Hubble constant~\cite{2010ApJ...725..496N}.

To this end, we have the ambition of reporting gravitational-wave candidates
not minutes \emph{after} the merger, as has already been accomplished, but
seconds \emph{before}.  By
looking for threshold crossings before the gravitational-wave signal leaves
the detection band, it is possible to trade some signal to noise ratio (\SNR{})
for latency.  Figure \ref{fig:earlywarning} shows projected early trigger rates
for \textsc{ns}--\textsc{ns} binaries in Advanced \LIGO\ assuming the event rate predictions
in~\cite{Abadie:2010p10836}.
%
\begin{figure}
\begin{center}
\includegraphics{figures/snr_in_time.pdf}
\caption{\label{fig:earlywarning} Expected number of \textsc{ns}--\textsc{ns} sources that will
be detectable $t$ seconds before coalescence.  The heavy solid line is the most
likely yearly rate estimate $\dot N_{\mathrm{re}}$ during Advanced \LIGO.  The
light solid lines represent the confidence interval $[\dot N_{\mathrm{low}}, \dot 
N_{\mathrm{high}}]$ described in
\cite{Abadie:2010p10836}.  Advanced \LIGO\ noise projections are described in 
\cite{ALIGONoise}.  Assuming that an \SNR\ of 8 is sufficient for
detection and that we observe $\dot N_{\mathrm{re}} = 40$ events~yr$^{-1}$ with a
detector in the `zero detuning, high power' configuration, about 10~yr$^{-1}$ will be 
detectable within 10 seconds before merger and $\sim1$~yr$^{-1}$ will be detectable 
within 100 seconds of merger.  The `zero detuning, lower power' and `\textsc{bbh} optimized' configurations are still more conducive to advance detection
but at the cost of fewer total detectable events above an \SNR\ of 8.}
\end{center}
\end{figure}

\editorial{Citation needed for LOOC-UP} The gravitational-wave community
initiated a project to send alerts when potential gravitational-wave transients
are observed.  In October 2010, \LIGO{} completed its sixth science run
(S6) and Virgo completed its third science run (VSR3).  While both
\LIGO{} detectors and Virgo were operating, several all-sky detection
pipelines operated in a low-latency configuration, namely \textsc{mbta}, ihope,
Coherent WaveBurst, and Omega~\cite{HugheyGWPAW2011, S6lowlatency}.
\editorial{Get references for these low-latency pipelines.} The S6 analyses
achieved latencies of 30--60 minutes, which were dominated by a human vetting
process. Candidates were sent for electromagnetic followup to several
telescopes; Swift, \textsc{rotse}, \textsc{tarot}, and Zadko~\cite{kanner2008,
HugheyGWPAW2011} took images of likely sky locations.  \textsc{mbta} achieved
the best gravitational-wave trigger generation latencies of 2--5 minutes.  We
assume that in the advanced detector era the vetting process will be automated,
so current gravitational wave search methodology and telescope actuation would
dominate latency.

Advance detection of compact binary coalescences (\CBC{}s) will require striking a balance between latency
and throughput.  \CBC{} searches consist of banks of matched filters, or
cross-correlations between the data stream and a bank of nominal ``template''
signals.  There are many different implementations of matched filters, but most
have high throughput at the cost of high latency, or low latency at the cost of
low throughput.  The former are epitomized by the overlap-save algorithm
\cite{numerical-recipes-chapter-13} for
frequency domain (\FD) convolution, currently the preferred method in gravitational wave
searches.  The most obvious example of the latter is the time domain
(\TD) convolution, which has no latency at all.  However, its
computational complexity is quadratic in the length of the templates, so it is
prohibitively expensive for long templates.

@@

The paper is organized as follows. First we provide an overview of our method
for detecting compact binary coalescence signals in an \earlywarning\ analysis.
We then describe the pipeline we have constructed that implements our method.
To validate the approach we present results of simulations and conclude with
some remarks on what remains to prepare for the Advanced detector era.

>>>>

As a compact binary system loses energy to gravitational waves, its orbital
separation decays, leading to a run-away inspiral with the
gravitational-wave amplitude and frequency increasing until the system
eventually merges soon after the innermost stable circular orbit (\ISCO). If a
neutron star is involved it may become tidally disrupted near the merger.  This
disrupted matter can fuel a bright electromagnetic counterpart in the merger's aftermath~\cite{shibata:2007}.
The coalescence of compact binary systems consisting of neutron stars
(\textsc{ns}) and/or black holes (\textsc{bh}) is the most promising source of
gravitational radiation for Advanced \LIGO, Virgo, \GEO\, and
\LCGT~\cite{ALIGOWeb, AVirgoWeb, GEOWeb, LCGTWeb}.  Tens of binary coalescence
events are expected to be observed in the advanced detector era later this
decade~\cite{Abadie:2010p10836}. In the era of first-generation
detectors the gravitational-wave community
initiated a project to send alerts when potential gravitational-wave transients
were observed in order to trigger followup observations by optical telescopes.
The best demonstrated latencies were 30--60 minutes~\cite{LOOCUP}, which was
an important achievement, but far too late to catch prompt electromagnetic emission.

Prompt electromagnetic emission can arise as shells of relativistically
out-flowing matter collide in the inner shock. Such an inner shock from a
compact binary coalescence is believed to be a mechanism for short gamma-ray
bursts (short \GRB{}s)~\cite{Lee:2005, nakar07}. The same inner shocks, or
potentially reverse shocks, can produce a bright accompanying optical
flash~\cite{Sari99}. Prompt emission is a probe into the extreme initial
conditions of the outflow, in contrast with afterglows, which arise in the
external shock with the local medium and are relatively
insensitive to initial conditions. Optical flashes have only been observed to
peak within tens of seconds for a handful of long
\GRB{}s~\cite{2011CRPhy..12..255A} by telescopes with extremely rapid response
or, in the case of \textsc{grb 080319b}, by pure serendipity, where several
telescopes were observing a previous \GRB{} in the same
field~\cite{2008Natur.455..183R}. Short \GRB{}s, on the other hand, typically
fade too quickly to even catch the tails of the afterglows. Rapid
gravitational-wave transient alerts could enable the first observation of
prompt optical flashes and the rise of afterglows from short \GRB{}s, and, for
an exceptional event, perhaps even a glimpse of a
precursor~\cite{0004-637X-723-2-1711}. An electromagnetic counterpart would
vastly boost confidence in the gravitational-wave detection and provide the
tight sky localization necessary to allow determination of the source's host
galaxy, which leads to a redshift measurement. A coincident gravitational-wave
observation will provide a luminosity distance measurement. With these
together, we can produce precision measurements of the Hubble
constant~\cite{2010ApJ...725..496N}.

To this end, we have the ambition of reporting gravitational-wave candidates
not minutes \emph{after} the merger, as has already been accomplished, but
seconds \emph{before}.  By examining the signal to noise ratio (\SNR{}) stream
for threshold crossings before the gravitational-wave signal leaves the
detection band, it is possible to trade some \SNR{}
for latency.  Figure \ref{fig:earlywarning} shows projected early trigger rates
for \textsc{ns}--\textsc{ns} binaries in Advanced \LIGO\ assuming the event rate predictions
in~\cite{Abadie:2010p10836}.
%
\begin{figure}
\begin{center}
\includegraphics{figures/snr_in_time.pdf}
\caption{\label{fig:earlywarning} Expected number of \textsc{ns}--\textsc{ns} sources that will
be detectable $t$ seconds before coalescence.  The heavy solid line is the most
likely yearly rate estimate $\dot N_{\mathrm{re}}$ during Advanced \LIGO.  The
light solid lines represent the confidence interval $[\dot N_{\mathrm{low}}, \dot 
N_{\mathrm{high}}]$ described in
\cite{Abadie:2010p10836}.  Advanced \LIGO\ noise projections are described in 
\cite{ALIGONoise}.  Assuming that an \SNR\ of 8 is sufficient for
detection and that we observe $\dot N_{\mathrm{re}} = 40$ events~yr$^{-1}$ with a
detector in the `zero detuning, high power' configuration, about 10~yr$^{-1}$ will be 
detectable within 10 seconds before merger and $\sim1$~yr$^{-1}$ will be detectable 
within 100 seconds of merger.  The `zero detuning, lower power' and `\textsc{bbh} optimized' configurations are still more conducive to advance detection
but at the cost of fewer total detectable events above an \SNR\ of 8.}
\end{center}
\end{figure}

In October 2010, \LIGO{} completed its sixth science run
(S6) and Virgo completed its third science run (VSR3).  While both
\LIGO{} detectors and Virgo were operating, several all-sky detection
pipelines operated in a low-latency configuration to send astronomical alerts,
namely \textsc{mbta}, ihope,
Coherent WaveBurst, and Omega~\cite{HugheyGWPAW2011, S6lowlatency}.
\editorial{Get references for these low-latency pipelines.} The S6 analyses
achieved latencies of 30--60 minutes, which were dominated by a human vetting
process. Candidates were sent for electromagnetic followup to several
telescopes; Swift, \textsc{rotse}, \textsc{tarot}, and Zadko~\cite{kanner2008,
HugheyGWPAW2011} took images of likely sky locations.  \textsc{mbta} achieved
the best gravitational-wave trigger-generation latencies of 2--5 minutes.  We
assume that in the advanced detector era the vetting process will be automated,
so current gravitational-wave search methodology and telescope actuation would
dominate latency.

Advance detection of compact binary coalescences (\CBC{}s) will require striking a balance between latency
and throughput.  \CBC{} searches consist of banks of matched filters, or
cross-correlations between the data stream and a bank of nominal ``template''
signals.  There are many different implementations of matched filters, but most
have high throughput at the cost of high latency, or low latency at the cost of
low throughput.  The former are epitomized by the overlap-save algorithm
\cite{numerical-recipes-chapter-13} for frequency domain (\FD) convolution,
currently the preferred method in gravitational-wave
searches.  The most obvious example of the latter is the time domain
(\TD) convolution, which has no algorithmic latency.  However, its
computational complexity is quadratic in the length of the templates, so it is
prohibitively expensive for long templates.

@@

The paper is organized as follows. First we provide an overview of our method
for detecting compact binary coalescence signals in an \earlywarning\ analysis.
We then describe the pipeline we have constructed that implements our method.
To validate the approach we present results of simulations and conclude with
some remarks on what remains to prepare for the advanced detector era.

### 2011-06-06 method.tex
lloid_paper: addressed comments from Steve and Missy; significant introduction reordering
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@944 ccdd05d5-56c1-461a-bf98-85f6876523b4

In this section we describe a decomposition of the \CBC{} waveform
space that reduces low-latency filtering cost sufficiently to allow for the
possibility of \earlywarning\ detection with modest computing requirements.  We
expand on the ideas of~\cite{Marion2004, Buskulic2010} that describe a
multiband decomposition of the compact binary parameter space that resulted in
a search with minutes latency in \LIGO{}'s S6 and Virgo's VSR2 science runs.
We combine this with the \SVD\ rank-reduction method described
in~\cite{Cannon:2010p10398} that exploits the redundancy of the template banks.

@@

\subsubsection{Extrapolation of the computational cost to an Advanced LIGO search}
\editorial{Drew: I think this section be the last section of the results section. Should we move it to be Sec. 4.2?}

@@

By comparison, using the \TD\ method to achieve the same latency costs $\sim 10^{13}$ \flops\ for this particular sub-bank, and so it would require $\sim 10^6$ present-day machines.  Presently, the \LIGO\ data grid consists of only $\sim 10^4$ machines.

>>>>

In this section we describe a decomposition of the \CBC{} waveform
space that reduces low-latency filtering cost sufficiently to allow for the
possibility of \earlywarning\ detection with modest computing requirements.  We
expand on the ideas of~\cite{Marion2004, Buskulic2010} that describe a
multiband decomposition of the compact binary parameter space that resulted in
a search with minutes latency in \LIGO{}'s S6 and Virgo's VSR3 joint science
runs.  We combine this with the \SVD\ rank-reduction method described
in~\cite{Cannon:2010p10398} that exploits the redundancy of the template banks.

@@

\subsubsection{Extrapolation of the computational cost to an Advanced \LIGO{} search}
\editorial{Drew: I think this section be the last section of the results section. Should we move it to be Sec. 4.2? Nick: I'm inclined to agree with Drew, though I see that we want to make explicit that it's an analysis of the algorithm and not of the implementation and the current placement aids that. Also, the given FLOPS numbers correspond to what choice of \SVD{} tolerance and $N^\shortuparrow$?}

@@

By comparison, using the \TD\ method to achieve the same latency costs $\sim 10^{13}$ \flops\ for this particular sub-bank, and so it would require $\sim 10^6$ present-day machines.  Presently, the \LIGO{} Data Grid consists of only $\sim 10^4$ machines.
